Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-10 ind_files/04_unification_discussion.tex
Minor phrasing changes

\eqref{nD_stochastic_RM} decompose into a selection term $\textrm{Cov}(\tau,f)$ for reduced turnover rates and a transmission bias term that vanishes in the weak mutation ($\mu \to 0$) limit. Following~\cite{constable_demographic_2016} and ~\cite{week_white_2021}, we refer to the effect of the covariance term (the second term of equation \eqref{nD_stochastic_Price}) as \emph{noise-induced selection} since it occurs exactly analogously to classical natural selection (but for lower $\tau$) and is induced purely by the finiteness of the population. Since this evolutionary force is unique to finite populations and has therefore been overlooked in classical population genetics, it warrants some more detailed discussion. Biologically, the $\textrm{Cov}(\tau,f)$ term (with a negative sign) describes a biasing effect due to differential turnover rates and can intuitively be understood as being similar to gambler's ruin in probability theory through the following reasoning: If a type $i$ has a higher $\tau_i$, it experiences greater turnover due to a generally higher birth and death rate and thus experience more births and deaths in a given time interval than an otherwise equivalent species with a lower $\tau_i$. More events mean greater demographic stochasticity, and types with a higher $\tau_i$ thus tend to be eliminated by a stochastic analog of selection because they experience more chance events (births and deaths) in a given time period. This effect is less visible if the total population size is higher because larger populations generally experience less stochasticity, which is reflected in the $1/N_K$ factor in this term. This stochastic analog of selection for reduced turnover rates, captured by the second term of equation \eqref{nD_stochastic_RM}, is the force responsible for the `reversal of the direction of deterministic selection' induced by demographic noise in previous studies~\citep{houchmandzadeh_selection_2012, houchmandzadeh_fluctuation_2015, constable_demographic_2016, mcleod_social_2019}.Note that types that tend to increase the \emph{total} population size $KN_K(t)$ (such as altruists in evolutionary theory and mutualists in ecological communities) will reduce the magnitude of this effect compared to types that do not facilitate such an increase, such as cheaters and highly competitive species. Further, if altruists/mutualists act by reducing the death rate (rather than increasing the birth rate) of other individuals, their presence causes higher $w$ and lower $\tau$ in the beneficiary individuals, both of which are favored by selection (but note that if they act by increasing the birth rate, they increase the magnitude of negative noise-induced selection disfavoring the beneficiary individual), which explains why this effect preferentially favors mutualists in reversing the direction of deterministic selection in finite population models with fluctuating population sizes if interaction effects are on the death rate~\citep{mcleod_social_2019}. Thus, selection for reduced turnover rate could help explain why cooperation often persists in finite population IbMs of social evolution~\citep{houchmandzadeh_selection_2012,houchmandzadeh_fluctuation_2015,chotibut_evolutionary_2015,behar_fluctuations-induced_2016,mcleod_social_2019} despite infinite population models predicting their extinction. The fact that total population size controls the strength of noise-induced selection also explains why cooperation is favored in the early transient period of population growth~\citep{melbinger_evolutionary_2010} when simulations are initiated from a small population size --- In the early transient period, $N_K(t)$ is small, and the biasing effect of differential turnover rates is stronger, thus favoring cooperation. The fact that the entire term scales inversely with the total population size $KN_K(t)$ suggests that the effect of this force is weak for large populations, which explains why the persistence of cooperators is often only observed in restrictive sounding conditions such as quasi-neutrality, timescale separation, or a weak selection + weak mutation limit~\citep{mcleod_social_2019}. In all three of these cases, the first term on the RHS of \eqref{nD_eqn_for_frequencies} becomes identically 0. It therefore no longer contributes to the trait frequency dynamics, thus allowing us to see the (otherwise weak) contributions of the second term.

The third term of \eqref{nD_stochastic_Price} is relevant in both finite and infinite populations whenever $f_i$ can vary over time and represents feedback effects on the quantity $f_i$ of the $i\textsuperscript{th}$ species over short (`ecological') time-scales. Such feedback is often through a changing environment or phenotypic/behavioral plasticity, but other biological phenomena may also be at play. The fourth term of \eqref{nD_stochastic_Price} is a transmission bias term, with a correction factor due to noise-induced selection. Finally, the last term of \eqref{nD_stochastic_Price} describes the role of stochastic fluctuations. The contributions of this last term are `directionless' due to the $dB_t$ factors, and this term vanishes when we take a conditional expectation value over the underlying probability space. We denote this probabilistic expectation value operation by $\mathbb{E}[\cdot]$ to distinguish it from the statistical mean \eqref{nD_mean}. Note that this expectation is conditioned on the initial state of the population, and thus $\mathbb{E}[\cdot]$ is really shorthand for $\mathbb{E}[\ \cdot \ | \ \mathbf{X}_0 = \mathbf{x}_0]$.

@@

To the best of my knowledge, a general formulation of stochastic field equations for the population dynamics of quantitative traits from the functional Fokker-Planck side in the manner I have carried out here has not been done before. Similar equations have been formulated for some specific stochastic models of quantitative trait evolution~\citep{rogers_demographic_2012,rogers_modes_2015} and population ecology of size-structured communities~\citep{odwyer_integrative_2009}. Stochastic field equations are also known in mathematical neurobiology~\citep{buice_field-theoretic_2007,bressloff_stochastic_2010,coombes_neural_2014}, and have recently been proposed in a model of collective motion~\citep{o_laighleis_minimal_2018}, suggesting that the study and formalization of such equations by pure mathematicians could have wide-spread applications. As is, such field equations are not on very solid mathematical footing, and are currently primarily used by physicists working in areas such as statistical field theory and are largely attacked using ingenious (and often heuristic) arguments and tools that are not necessarily mathematically well understood~\citep{carmona_stochastic_1999}. Equation \eqref{disc_functional_field_eqns} opens up the study of quantitative trait dynamics in finite fluctuating populations to analysis using various powerful tools of physics and stochastic processes such as duality~\citep{greenman_duality_2020}, the Fock space representation~\citep{dodd_many-body_2009, del_razo_probabilistic_2022}, and the path integral formalism~\citep{doi_second_1976, peliti_path_1985, dodd_many-body_2009, chow_path_2015, weber_master_2017}. Equation \eqref{disc_functional_field_eqns} also provides a more accessible alternative formulation that does not (explicitly) rely on tools from relatively `advanced' mathematical fields such as martingale theory and measure-valued processes that are typically used to study the evolution of quantitative traits in finite, fluctuating populations~\citep{dawson_stochastic_1975,fleming_measure-valued_1979,ethier_markov_1986,champagnat_unifying_2006,etheridge_mathematical_2011,week_white_2021}, thus hopefully making the field more accessible to theorists without formal training in mathematical areas such as measure theory and functional analysis.

@@

Despite these practical impediments to actually solving equations for particular models, my equations are very general, since part \ref{part_theory} makes essentially no assumptions other than density dependence, the impossibility of infinite growth starting from finite population size, and the ability to define per-capita birth and death rates. Further, the terms of the equations I derive always lend themselves to simple biological interpretation. Thus, like the Price equation, the utility of these equations lies not in their solutions for specific models, but instead in their generality and the fact that their terms help us clearly understand the various forces operating in biological populations~\citep{frank_natural_2012,luque_one_2017, luque_mirror_2021}. The general approach of working with `high-level' processes without specifying system-specific details is part of a broader pursuit of `model-independent' eco-evolutionary theory that has recently been gaining popularity in the literature ~\citep{grafen_formal_2014, queller_fundamental_2017, doebeli_towards_2017, lion_theoretical_2018, allen_mathematical_2019, rice_universal_2020, week_white_2021, wickman_theoretical_2022, kuosmanen_turnover_2022, mazzolini_universality_2022,lion_extending_2023}. Such model-independent descriptions that `abstract away' system-specific details almost inevitably come at the cost of precision~\citep{levins_strategy_1966, potochnik_idealization_2018}, and are thus intended to complement rather than supersede the study of specific models of specific systems.

>>>>

\eqref{nD_stochastic_RM} decompose into a selection term $\textrm{Cov}(\tau,f)$ for reduced turnover rates and a transmission bias term that vanishes in the weak mutation ($\mu \to 0$) limit. Following~\cite{constable_demographic_2016} and ~\cite{week_white_2021}, we refer to the effect of the covariance term (the second term of equation \eqref{nD_stochastic_Price}) as \emph{noise-induced selection} since it occurs exactly analogously to classical natural selection (but for lower $\tau$) and is induced purely by the finiteness of the population. Since this evolutionary force is unique to finite populations and has therefore been overlooked in classical population genetics, it warrants some more detailed discussion. Biologically, the $\textrm{Cov}(\tau,f)$ term (with a negative sign) describes a biasing effect due to differential turnover rates and can intuitively be understood as being similar to gambler's ruin in probability theory through the following reasoning: If a type $i$ has a higher $\tau_i$, it experiences greater turnover due to a generally higher birth and death rate and thus experience more births and deaths in a given time interval than an otherwise equivalent species with a lower $\tau_i$. More events mean greater demographic stochasticity, and types with a higher $\tau_i$ thus tend to be eliminated by a stochastic analog of selection because they experience more chance events (births and deaths) in a given time period. This effect is less visible if the total population size is higher because larger populations generally experience less stochasticity, which is reflected in the $1/N_K$ factor in this term. This stochastic analog of selection for reduced turnover rates, captured by the second term of equation \eqref{nD_stochastic_RM}, is the force responsible for the `reversal of the direction of deterministic selection' induced by demographic noise in previous studies~\citep{houchmandzadeh_selection_2012, houchmandzadeh_fluctuation_2015, constable_demographic_2016, mcleod_social_2019}. Note that types that tend to increase the \emph{total} population size $KN_K(t)$ (such as altruists in evolutionary theory and mutualists in ecological communities) will reduce the magnitude of this effect compared to types that do not facilitate such an increase, such as cheaters and highly competitive species. Further, if altruists/mutualists act by reducing the death rate (rather than increasing the birth rate) of other individuals, their presence causes higher $w$ and lower $\tau$ in the beneficiary individuals, both of which are favored by selection (but note that if they act by increasing the birth rate, they increase the magnitude of negative noise-induced selection disfavoring the beneficiary individual), which explains why this effect preferentially favors mutualists in reversing the direction of deterministic selection in finite population models with fluctuating population sizes if interaction effects are on the death rate~\citep{mcleod_social_2019}. Thus, selection for reduced turnover rate could help explain why cooperation often persists in finite population IbMs of social evolution~\citep{houchmandzadeh_selection_2012,houchmandzadeh_fluctuation_2015,chotibut_evolutionary_2015,behar_fluctuations-induced_2016,mcleod_social_2019} despite infinite population models predicting their extinction. The fact that total population size controls the strength of noise-induced selection also explains why cooperation is favored in the early transient period of population growth~\citep{melbinger_evolutionary_2010} when simulations are initiated from a small population size --- In the early transient period, $N_K(t)$ is small, and the biasing effect of differential turnover rates is stronger, thus favoring cooperation. The fact that the entire term scales inversely with the total population size $KN_K(t)$ suggests that the effect of this force is weak for large populations, which explains why the persistence of cooperators is often only observed in restrictive sounding conditions such as quasi-neutrality, timescale separation, or a weak selection + weak mutation limit~\citep{mcleod_social_2019}. In all three of these cases, the first term on the RHS of \eqref{nD_eqn_for_frequencies} becomes identically 0. It therefore no longer contributes to the trait frequency dynamics, thus allowing us to see the (otherwise weak) contributions of the second term.

The third term of \eqref{nD_stochastic_Price} is relevant in both finite and infinite populations whenever $f_i$ can vary over time and represents feedback effects on the quantity $f_i$ of the $i\textsuperscript{th}$ species over short (`ecological') time-scales. Such feedback is often through a changing environment or phenotypic/behavioral plasticity, but other biological phenomena may also be at play. The fourth term of \eqref{nD_stochastic_Price} is a transmission bias term, with a correction factor due to noise-induced selection. Finally, the last term of \eqref{nD_stochastic_Price} describes the role of stochastic fluctuations. The contributions of this last term are `directionless' due to the $dB_t$ factors, and this term vanishes when we take a conditional expectation value over the underlying probability space. I will denote this probabilistic expectation value operation by $\mathbb{E}[\cdot]$ to distinguish it from the statistical mean $\overline{f}$ of a type-level quantity in the population (see section \ref{sec_stat_measures} for a brief discussion on the difference between statistical averages and probabilistic expectations in this context). Note that this expectation is conditioned on the initial state of the population, and thus $\mathbb{E}[\cdot]$ is really shorthand for $\mathbb{E}[\ \cdot \ | \ \mathbf{X}_0 = \mathbf{x}_0]$.

@@

To the best of my knowledge, a general formulation of stochastic field equations for the population dynamics of quantitative traits from the functional Fokker-Planck side in the manner I have carried out here has not been done before. Similar equations have been formulated for some specific stochastic models of quantitative trait evolution~\citep{rogers_demographic_2012,rogers_modes_2015} and population ecology of size-structured communities~\citep{odwyer_integrative_2009}. Stochastic field equations are also known in mathematical neurobiology~\citep{buice_field-theoretic_2007,bressloff_stochastic_2010,coombes_neural_2014}, and have recently been proposed in a model of collective motion~\citep{o_laighleis_minimal_2018}, suggesting that the study and formalization of such equations by pure mathematicians could have wide-spread applications. As is, such field equations are not on very solid mathematical footing, and are currently primarily used by physicists working in areas such as statistical field theory and are largely attacked using ingenious (and often heuristic) arguments and tools that are not necessarily mathematically well understood~\citep{carmona_stochastic_1999}. Equation \eqref{disc_functional_field_eqns} opens up the study of quantitative trait dynamics in finite fluctuating populations to analysis using various powerful tools of physics and stochastic processes such as duality~\citep{greenman_duality_2020}, the Fock space representation~\citep{dodd_many-body_2009, del_razo_probabilistic_2022}, and the path integral formalism~\citep{doi_second_1976, peliti_path_1985, dodd_many-body_2009, chow_path_2015, weber_master_2017}. Equation \eqref{disc_functional_field_eqns} also provides a more accessible alternative formulation that does not (explicitly) rely on tools from relatively `advanced' mathematical fields such as martingale theory and measure-valued processes that are typically used to study the evolution of quantitative traits in finite, fluctuating populations~\citep{dawson_stochastic_1975,fleming_measure-valued_1979,ethier_markov_1986,champagnat_unifying_2006,etheridge_mathematical_2011,week_white_2021}. This is intended to make the models and ideas more accessible to theorists without formal training in mathematical areas such as measure theory and functional analysis.

@@

Despite these practical impediments to actually solving equations for particular models, my equations are very general, since part \ref{part_theory} makes essentially no assumptions other than density dependence, the impossibility of infinite growth starting from finite population size, and the ability to define per-capita birth and death rates. Further, the terms of the equations I derive always lend themselves to simple biological interpretation. Thus, like the Price equation, the utility of these equations lies not in their solutions for specific models, but instead in their generality and the fact that their terms help us clearly understand the various forces operating in biological populations~\citep{frank_natural_2012,luque_one_2017, luque_mirror_2021}. Model-independent descriptions such as this `abstract away' system-specific details and almost inevitably come at the cost of precision, and are thus intended to complement rather than supersede specific models of specific systems~\citep{levins_strategy_1966, potochnik_idealization_2018}. The general spirit of this thesis is, however, part of a broader pursuit of `model-independent' eco-evolutionary theory that has recently been gaining popularity in the literature~\citep{grafen_formal_2014, queller_fundamental_2017, doebeli_towards_2017, lion_theoretical_2018, allen_mathematical_2019, rice_universal_2020, week_white_2021, wickman_theoretical_2022, kuosmanen_turnover_2022, mazzolini_universality_2022,lion_extending_2023,allen_natural_2023}.
